---
title: "Example code for dimensionality reduction and clustering in R"
author: "Kevin Rue-Albrecht"
date: "03/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(cowplot)
library(umap)
library(Rtsne)
library(dendextend)
library(dbscan)
```

# Exercise

## Setup

- Import the `iris` data set.

```{r}

iris

```

- Separate the matrix of measurements in a new object named `iris_features`.

```{r}
iris_features <- as.matrix(iris[, 1:4])
head(iris_features)
```

# Exercise

## Apply Principal Components Analysis (PCA)

The `prcomp()` function allows you to standardise the data as part of the principal components analysis itself.

- Apply PCA while centering and scaling the matrix of features.

```{r}
##Because a sepal is typically much smaller than a petal, scaling and centering is important
pca_iris <- prcomp(iris_features, center = TRUE, scale. = TRUE)
pca_iris
```

- Examine the PCA output.
  Display the loading of each feature on each principal component.

```{r}

pca_iris$rotation

##This tells us the PC values for each of the variables

```

```{r}
pca_iris$x

##this tells us the PC values for each of the samples (in this case flowers)
```

- Use the return value of the PCA to create a `data.frame` called `pca_iris_dataframe` that contains the coordinates projected on principal components.

```{r}
pca_iris_dataframe <- as.data.frame(pca_iris$x)
head(pca_iris_dataframe)

summary(pca_iris)
```

- Visualise the PCA projection using `ggplot2::geom_point()`.

```{r}
ggplot(pca_iris_dataframe, aes(x = PC1, y = PC2)) +
  geom_point()
  
  
```

### Bonus point

- Color data points according to their class label.

- Store the PCA plot as an object named `pca_iris_species`.

```{r}

head(pca_iris_dataframe)
speciespca_iris_dataframe <- cbind(pca_iris_dataframe, iris$Species)
speciespca_iris_dataframe


ggplot(speciespca_iris_dataframe, aes(x = PC1, y = PC2)) +
  geom_point(aes(colour = iris$Species))


```

```{r}
pca_iris_species <- ggplot(speciespca_iris_dataframe, aes(x = PC1, y = PC2)) +
  geom_point(aes(colour = iris$Species))
  
  
pca_iris_species
```

# Exercise

## Variable loading

- Color a scatter plot of PC1 and PC2 by the value of the variable most strongly associated with the first principal component.

What do you observe?

```{r}

head(pca_iris_dataframe)

head(pca_iris$rotation)
```

```{r}

petal_length <
ggplot(speciespca_iris_dataframe, aes(x = PC1, y = PC2)) +
  geom_point(aes(colour = iris$Petal.Length))
  
  

```

> Answer:
> 
> 

## Variance explained

- Compute the variance explained by principal components, using information present in the return value of the `prcomp()` function.

```{r}
str(pca_iris)

explained_variance_ratio <- pca_iris$sdev^2/ sum(pca_iris$sdev^2)
explained_variance_ratio
```

- Visualise the variance explained by each principal component using `ggplot2::geom_col()`.

```{r}
variance_dataframe <- cbind(explained_variance_ratio, c("PC1", "PC2", "PC3", "PC4"))
variance_dataframe <- as.data.frame(variance_dataframe)
head(variance_dataframe)
variance_dataframe
```

```{r}
ggplot(variance_dataframe, aes(x = V2, y = explained_variance_ratio)) +
         geom_col()

  
  
```

# Exercise

## UMAP

- Apply UMAP on the output of the PCA.

```{r}
set.seed(1) # Set a seed for reproducible results
umap_iris <- umap(pca_iris$x)
umap_iris
```

- Inspect the UMAP output.

```{r}
summary(umap_iris)

str(umap_iris)
```

- Visualise the UMAP projection using `ggplot2::geom_point()`.

```{r}
umap_iris_dataframe <- as.data.frame(umap_iris$layout)

head(umap_iris_dataframe)
```

```{r}
ggplot(umap_iris_dataframe, aes(x = V1, y = V2)) +
  geom_point()
  
  
```

### Bonus point

- Color data points according to their class label.

- Store the UMAP plot as an object named `umap_iris_species`.

```{r}

head(umap_iris_dataframe) 

```

```{r}
umap_iris_species <- ggplot(umap_iris_dataframe, aes(x = V1, y = V2)) +
  geom_point(aes(colour = iris$Species))
  
  
umap_iris_species
```

# Exercise

## t-SNE

- Apply t-SNE and inspect the output.

```{r}
set.seed(1) # Set a seed for reproducible results
?Rtsne
tsne_iris <- Rtsne(pca_iris$x, dims = 2, pca = FALSE, check_duplicates = FALSE)
str(tsne_iris)
```

- Use the return value of the t-SNE to create a `data.frame` called `tsne_iris_dataframe` that contains the coordinates.

```{r}
tsne_iris_dataframe <- as.data.frame(tsne_iris$Y)

head(tsne_iris_dataframe)
```

- Visualise the t-SNE projection.

```{r}
ggplot(tsne_iris_dataframe, aes(x = V1, y = V2) ) +
  geom_point()
  
  
```

### Bonus points

- Color data points according to their class label.

- Store the t-SNE plot as an object named `tsne_iris_species`.

```{r}

head(tsne_iris_dataframe)
```

```{r}
tsne_iris_species <- ggplot(tsne_iris_dataframe, aes(x = V1, y = V2) ) +
  geom_point(aes(colour = iris$Species))
  
  
tsne_iris_species
```

- Combine PCA, UMAP and t-SNE plots in a single figure.

```{r, fig.height=3, fig.width=10}
cowplot::plot_grid(pca_iris_species, umap_iris_species, tsne_iris_species, nrow = 1)
```

# Exercise

## Hierarchical clustering

- Perform hierarchical clustering on the `iris_features` data set,
  using the `euclidean` distance and method `ward.D2`.
  Use the functions `dist()` and `hclust()`.

```{r}
?dist
dist_iris <- dist(iris_features, method = "euclidean")


?hclust
hclust_iris_ward <- hclust(dist_iris, method = "ward.D2")
hclust_iris_ward
```

- Plot the clustering tree.
  Use the function `plot()`.

```{r}
plot(hclust_iris_ward)

##Height is generally the distance between the clusters
```

How many clusters would you call from a visual inspection of the tree?

> Answer:
> Between 3 and 6 
> 

- **Bonus point:** Color leaves by known species (use `dendextend`).

```{r}
?dendextend
iris_hclust_dend <- as.dendrogram(hclust_iris_ward)
labels_colors(iris_hclust_dend) <- as.numeric(iris$Species) ##because for this graph to see colour it needs to be numeric 
plot(iris_hclust_dend)
```

- Cut the tree in 3 clusters and extract the cluster label for each flower.
  Use the function `cutree()`.

```{r}
?cutree
iris_hclust_ward_labels <- cutree(hclust_iris_ward, k = 3)
iris_hclust_ward_labels


```

- Repeat clustering using 3 other agglomeration methods:

  + `complete`
  + `average`
  + `single`

```{r}

##hclust_iris_ward <- hclust(dist_iris, method = "ward.D2")
# complete
hclust_iris_complete <- hclust(dist_iris, method = "complete")
iris_hclust_complete_labels <- cutree(hclust_iris_complete, k = 3)
iris_hclust_complete_labels
```

```{r}
# average
hclust_iris_average <- hclust(dist_iris, method = "average")
iris_hclust_average_labels <- cutree(hclust_iris_average, k = 3)
iris_hclust_average_labels
```

```{r}
# single
hclust_iris_single <- hclust(dist_iris, method = "single")
iris_hclust_single_labels <- cutree(hclust_iris_single, k = 3)
iris_hclust_single_labels
```

- Compare clustering results on scatter plots of the data.

```{r}
iris_clusters_dataframe <- iris
iris_clusters_dataframe$hclust_average <- as.factor(iris_hclust_average_labels)
iris_clusters_dataframe$hclust_complete <- as.factor(iris_hclust_complete_labels)
iris_clusters_dataframe$hclust_single <- as.factor(iris_hclust_single_labels)
iris_clusters_dataframe$hclust_ward <- as.factor(iris_hclust_ward_labels)
```

```{r, fig.height=3, fig.width=7}
head(iris_clusters_dataframe)
plot_average <- ggplot(iris_clusters_dataframe, aes(x=Petal.Length, y=Petal.Width)) +
  geom_point(aes(colour = hclust_average))
  
  
plot_complete <- ggplot(iris_clusters_dataframe, aes(x=Petal.Length, y=Petal.Width)) +
  geom_point(aes(colour = hclust_complete))
  
  
plot_single <- ggplot(iris_clusters_dataframe, aes(x=Petal.Length, y=Petal.Width)) +
  geom_point(aes(colour = hclust_single))
  
  
plot_ward <- ggplot(iris_clusters_dataframe, aes(x=Petal.Length, y=Petal.Width)) +
  geom_point(aes(colour = hclust_average))
  
  
cowplot::plot_grid(plot_average, plot_complete, plot_single, plot_ward, nrow = 2)
 
  

```

# Exercise

## dbscan

- Apply `dbscan` to the `iris_features` data set.

```{r}
?dbscan
dbscan_iris <- dbscan(iris_features, eps = 0.9)
dbscan_iris
```

- Visualise the `dbscan` cluster label on a scatter plot of the data.

```{r}
iris_clusters_dataframe$dbscan <- as.factor(dbscan_iris)
head(iris_clusters_dataframe)
```

```{r}
dbscan_plot <- ggplot(iris_clusters_dataframe,    ) +
  
  
dbscan_plot
```

## hdbscan

- Apply `hdbscan` to the `iris_features` data set.

```{r}
hdbscan_iris <- hdbscan(    )
hdbscan_iris
```

- Visualise the `hdbscan` cluster label on a scatter plot of the data.

```{r}
iris_clusters_dataframe$hdbscan <- as.factor(   )
head(iris_clusters_dataframe)
```

```{r}
hdbscan_plot <- ggplot(iris_clusters_dataframe,    ) +
  
  
hdbscan_plot
```

## Bonus point

- Combine the plots of `dbscan` and `hdbscan` into a single plot.

```{r, fig.height=3, fig.width=6}
cowplot::plot_grid(
  
  
  
)
```

# Exercise

## K-means clustering

- Apply $K$-means clustering with $K$ set to 3 clusters.

```{r}
set.seed(1) # Set a seed for reproducible results
kmeans_iris <- kmeans(   )
kmeans_iris
```

- Inspect the output.

```{r}

```

- Extract the cluster labels.

```{r}

```

- Extract the coordinates of the cluster centers.

```{r}

```

- Construct a data frame that combines the `iris` dataset and the cluster label.

```{r}
iris_labelled <- iris
iris_labelled$Kmeans <- as.factor(   )
head(iris_labelled)
```

- Plot the data set as a scatter plot.

  + Color by cluster label.

```{r}
ggplot(iris_labelled,    ) +
  
  
```

### Bonus point

- Add cluster centers as points in the plot.

```{r}
iris_means_centers <- as.data.frame(   )
iris_means_centers$Kmeans <- as.factor(   )
head(iris_means_centers)
```


```{r}
ggplot(iris_labelled,    ) +
  
  
  
```

# Exercise

## Cross-tabulation with ground truth

- Cross-tabulate cluster labels with known labels.

```{r}
table(   )
```

How many observations are mis-classified by $K$-means clustering?

> Answer:
> 
> 
> 
> 
> 

## Elbow plot

- Plot the "total within-cluster sum of squares" for K ranging from 2 to 10.

```{r}

```

```{r}
get_mean_totss_for_k <- function(k, data) {
  kmeans_out <- kmeans(data, k)
  return(kmeans_out$tot.withinss)
}
k_range <- 2:10
kmean_totwithinss <- vapply(   )
kmean_totwithinss
```

```{r}
kmean_totwithinss_dataframe <- data.frame(
  K = ,
  totss = 
)
head(kmean_totwithinss_dataframe)
```

```{r}
ggplot(kmean_totwithinss_dataframe,    ) +
  
  
  
```

Do you agree that 3 is the optimal number of clusters for this data set?

> Answer:
> 
> 
> 
> 

