---
title: "Example code for single-cell droplet quality control"
author: "Kevin Rue-Albrecht"
date: "05/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DropletUtils)
library(tidyverse)
library(ggplot2)
library(cowplot)
library(scDblFinder)
library(Seurat)
library(DoubletFinder)
library(celda)
library(scater)
```

# Exercise

## Empty droplets

- Import the raw matrix into R; use `DropletUtils`.
  Make a `SingleCellExperiment` object called `sce`.

```{r}
library(DropletUtils)
sce <- read10xCounts(samples = c("pbmc5k"= "/project/obds/shared/resources/4_r_single_cell/singlecell_droplets/raw_feature_bc_matrix"), col.names = TRUE) ##need to include col.names = TRUE because the default is false
  
  

sce
```

- Remove droplets without any count at all.

This will massively speed up downstream steps, while removing unquestionable empty droplets.

```{r}
## assay(sce, "counts")
## we want to get rid of all the columns that have a sum of zero

sce <- sce[, colSums(assay(sce, "counts")) >0 ]
sce
```

- Run `DropletUtils::emptyDrops()`.

```{r}
set.seed(1)

##Important here you are not assigning this set back to the sce object, you are defining it as another variable called out 


out <- emptyDrops(sce, assay.type = "counts", lower = 100, niters = 10000) ##the lower arguement tells it not to calculate it if the count is less than 100, thus we get lots of NAs in the table generated
out

##the FDR column is the false discovery rate. Your FDR is the p value x by the number of tests 
```

**How do you read the output? Which droplets would you consider empty?**

> Answer:
> 
>
>
> 

- Remove empty droplets from the data sets.
  Call the new object `sce_filtered_emptydrops`.

```{r}

##we went to filter based on the results from the previous table - we want FDR<0.01, we are going to use which as well for the boolean

?which

positions <-which(out$FDR<0.01) ##the which here tells us it to return the position of where FDR is <0.01, then we store this in the variable positions which we can then use to subset the sce object


sce_filtered_emptydrops <- sce[, positions]
sce_filtered_emptydrops
```

## Bonus point

- Draw plots comparing empty droplets and other droplets
  (e.g., library size, mitochondrial content, PCA).

```{r}
##we need to create a table that combines some info from the sce and the out objects 
?tibble
plot_data <- tibble(library_size = colSums(assay(sce, "counts")), droplet_status = (out$FDR<0.01))
plot_data
  
  
ggplot(plot_data, aes(x = droplet_status, y = log(library_size))) +
  geom_violin()
  
##Alternatively 
ggplot(plot_data, aes(x = droplet_status, y = library_size)) +
  geom_violin() + scale_y_log10()
  

  
```

```{r}
is.mt <- 
# table(is.mt)
plot_data <- tibble(
  
  
  
  
)
ggplot(   ) +
  
  
  
```

```{r}

sce <- 
sce <- 
plot_data <- tibble(
  
  
)
ggplot(   ) +
  
  
```

- Import the filtered matrix for the same cells and compare their set of filtered cells to yours.

```{r}
sce_filtered_cellranger <- 
  
  


all_barcodes <- colnames(sce)

compare_table <- tibble(
  emptydrops = all_barcodes %in%    ,
  cellranger = all_barcodes %in%    ,
)
table(compare_table)
```

# Exercise

## Doublets

Continue using the data set after you excluded empty droplets.

- Run `scDblFinder::scDblFinder()`.

```{r}
?scDblFinder
set.seed(1)
sce_scdblfinder <- scDblFinder(sce_filtered_emptydrops)


sce_scdblfinder
colData(sce_scdblfinder)
table(sce_scdblfinder$scDblFinder.class)
```

**How do you read the output? Which droplets would you consider doublets?**

> Answer:
> 
> 
> 
> 

- Remove doublets from the data set.
  Call the new object `sce_filtered_scdblfinder`.

```{r}
is_singlet <- sce_scdblfinder$scDblFinder.class == "singlet"

sce_filtered_scdblfinder <- sce_scdblfinder[, is_singlet]
sce_filtered_scdblfinder
```

## Bonus point

- Draw plots comparing doublets and other droplets.
  (e.g., library size, mitochondrial content, PCA)

```{r}
##this is to do look at library size

?tibble 
plot_data_lib <- tibble(barcode = colnames(sce_scdblfinder),  library_size = colSums(counts(sce_scdblfinder)),                               doublet_type = sce_scdblfinder$scDblFinder.class)
  
  
ggplot(plot_data_lib, aes(x = doublet_type, y = log(library_size))) +
  geom_violin()
  
```

  
```{r}
is.mt <- 

plot_data <- tibble(
  
  
  
  
)
ggplot(   ) +
  
  
  
```

```{r}
sce_scdblfinder <- 
sce_scdblfinder <- 
plot_data <- tibble(
  
  
)

ggplot(   ) +
  
  
```

- Compare results with the output of `chris-mcginnis-ucsf/DoubletFinder`

  + You will need to make a `Seurat` object.

```{r, message=FALSE}
## Pre-process Seurat object (standard)
seu <- 
seu <- 
seu <- 
seu <- 
seu <- 
seu <- 
```

```{r, message=FALSE}
## pK Identification (no ground-truth) 

sweep.res.list <- 

sweep.stats <- 
bcmvn <- 
```

```{r}
ggplot() +
  
  
  
```

```{r}
## Homotypic Doublet Proportion Estimate 
homotypic.prop <- 
nExp_poi <- 
nExp_poi.adj <- 
```

```{r, message=FALSE}
## Run DoubletFinder with varying classification stringencies 
seu <- 
```

```{r}
table(   )
```

```{r, message=FALSE}
seu <- 
```

```{r}
table(   )
```

```{r}
doublets_table <- data.frame(
  
  
  
)
table(doublets_table)
```

# Exercise

## Ambient RNA

- Run `celda::decontX()` (continue using the data after excluding empty droplets and doublets).

```{r, message=FALSE}
?decontX
set.seed(1)
sce_decontX <- decontX(sce_filtered_scdblfinder, assayName = "counts")
```

- Visualize the UMAP layout created by `decontX`; use `plotDimReduceCluster()`.

```{r}
sce_decontX
umap <-
plotDimReduceCluster(sce_decontX$decontX_clusters, 
dim1 = reducedDim(sce_decontX, "decontX_UMAP")[,1], 
dim2 = reducedDim(sce_decontX, "decontX_UMAP")[,2],)
umap

```

- Visualize the percentage of contamination estimated by `decontX` in each cell.
  Use the `plotDecontXContamination()` function.

```{r}
?plotDecontXContamination
plotDecontXContamination(sce_decontX)
```

**Does any cluster display higher (estimated) levels of ambient RNA?**

> Answer:
> 
> 

- Visualize the log-normalised expression of marker genes on a UMAP layout faceted by gene; use `plotDimReduceFeature()`.

**Note:** For instance, use `CD3D`, `CD3E`, `GNLY`, `LYZ`, `S100A8`, `S100A9`, `CD79A`, `CD79B`, and `MS4A1`.

<!--
Source: https://bioconductor.org/packages/devel/bioc/vignettes/celda/inst/doc/decontX.pdf
(Section 5.3)
We will use:
- CD3D and CD3E for T-cells,
- LYZ, S100A8, and S100A9 for monocytes,
- CD79A, CD79B, and MS4A1 for B-cells,
- GNLY for NK-cells,
- and PPBP for megakaryocytes
-->

```{r}
## "CD3D", "CD3E", "GNLY", "LYZ", "S100A8", "S100A9", "CD79A", "CD79B", "MS4A1"
rowData(sce_decontX)
rownames(sce_decontX) <- uniquifyFeatureNames(rowData(sce_decontX)$ID, rowData(sce_decontX)$Symbol)
  


plotDimReduceFeature(counts(sce_decontX),
                     features = c("CD3D", "CD3E", "GNLY", "LYZ", "S100A8", "S100A9", "CD79A", "CD79B", "MS4A1", "PPBP"), 
                     dim1 = reducedDim(sce_decontX, "decontX_UMAP")[,1], dim2 = reducedDim(sce_decontX, "decontX_UMAP")[,2])
  
  

```

- Visualise the percentage of cells with detectable expression of marker genes before and after decontamination as a bar plot; use `plotDecontXMarkerPercentage()`.
We will use:
- CD3D and CD3E for T-cells,
- LYZ, S100A8, and S100A9 for monocytes,
- CD79A, CD79B, and MS4A1 for B-cells,
- GNLY for NK-cells,
- and PPBP for megakaryocytes

```{r}
markers <- list(Tcell = c("CD3E", "CD3D"), Bcell = c("CD79A", "CD79B", "MS4A1"), monocytes = c("LYZ", "S100A8", "S100A9"), NKcells = c("GNLY"), megakaryocytes = "PPBP")
  
  

cellTypeMappings <- list(Tcells = 1, NKcells = 5, monocytes = 2, Bcells = 3, megakaryoctyes = 7)
  
  
  
  
?plotDecontXMarkerPercentage
plotDecontXMarkerPercentage(sce_decontX, markers = markers, groupClusters = cellTypeMappings)
                        
  
  
  
  
)
```

```{r}
?plotDecontXMarkerPercentage
plotDecontXMarkerPercentage(
  sce_decontX,
  
  
  
)
```

```{r}
plotDecontXMarkerPercentage(
  
  
  
  
)
```

- Visualize the log-normalised expression of marker genes before and after decontamination as a violin plot; use `plotDecontXMarkerExpression()`.

```{r, fig.width=10, fig.height=10}
?plotDecontXMarkerExpression


plotDecontXMarkerExpression(sce_decontX, markers = markers, groupClusters = cellTypeMappings)
                       
  
##we havent done the normalisation yet so need to do that now

sce_decontX <- logNormCounts(sce_decontX, assay.type = "decontXcounts", name = "log_normalised_decontX_counts")
assays(sce_decontX)

sce_decontX <- logNormCounts(sce_decontX, assay.type = "counts", name = "log_normalised_counts")
assays(sce_decontX)

plotDecontXMarkerExpression(sce_decontX, markers = markers, groupClusters = cellTypeMappings)

```
